The Diagnostic Potential \& Interactive Dynamics of the Colorectal Cancer Virome

### 2018-10-11 doc/manuscript.tex
Doc updates including word doc

To identify the altered virus communities associated with colorectal
cancer, we built and tested random forest models for classifying stool
samples as belonging to individuals with either cancerous or healthy
colons. We confirmed that our bacterial 16S rRNA gene model replicated
the performance of the original report which used logit models instead
of random forest models \textbf{(Figure \ref{predmodel} A)} (8). We then
compared the bacterial OTU model to a model built using OVU relative
abundances. The viral model performed as well as the bacterial model
(corrected p-value = 0.4), with the viral and bacterial models achieving
mean area under the curve (AUC) values of 0.792 and 0.809, respectively
\textbf{(Figure \ref{predmodel} A - B)}. To evaluate the ability of both
bacterial and viral biomarkers to classify samples, we built a combined
model that used both bacterial and viral community data. The combined
model did not yield a statistically significant performance improvement
beyond the viral (corrected p-value = 0.4) and bacterial (corrected
p-value = 0.08) models, yielding an AUC of 0.768 \textbf{(Figure
\ref{predmodel} A - B)}.

We compared viral metagenomic methods to bacterial metagenomic methods
by building a viral model and a model built using OGU relative abundance
profiles from bacterial metagenomic shotgun sequencing data. This
bacterial model performed worse than the other models (mean AUC = 0.474)
\textbf{(Figure \ref{predmodel} A - B)}. To determine the cause of the
discrepancy between the two bacterial sequencing methods, we attempted
to compare the approaches at a common sequencing depth. This revealed
that the bacterial 16S rRNA gene model was strongly driven by sparse and
low abundance OTUs \textbf{(Figure \ref{16scompare})}. Removal of OTUs
with a median abundance of zero resulted in the removal of six OTUs, and
a loss of model performance down to what was observed in the
metagenome-based model \textbf{(Figure \ref{16scompare} A)}. The
majority of these OTUs had a relative abundance lower than 1\% across
the samples \textbf{(Figure \ref{16scompare} B)}. Although the features
in the viral model also were of low abundance \textbf{(Figure
\ref{threewaymodel} F)}, the coverage was sufficient for high model
performance, likely because viral genomes are orders of magnitude
smaller than bacterial genomes.

The association between the bacterial and viral communities and
colorectal cancer was driven by a few important microbes.
\emph{Fusobacterium} was the primary driver of the bacterial association
with colorectal cancer, which is consistent with its previously
described oncogenic potential \textbf{(Figure \ref{predmodel} C)}(22).
The virome signature also was driven by a few OVUs, suggesting a role
for these viruses in tumorigenesis \textbf{(Figure \ref{predmodel} D)}.
It is also important to note that while these viruses were driving the
signature, the magnitude of their importance and the significance of
those values was noticibly less than the bacterial 16S signature,
suggesting that unlike what is observed in the bacteria, there are many
viruses that are associated with the cancerous state. The identified
viruses were bacteriophages, belonging to \emph{Siphoviridae},
\emph{Myoviridae}, and phage taxa that could not be confidently
identified beyond their broad phage identification (i.e.
``unclassified''). Many of the important viruses, that were confirmed to
not have genomic similarity to known bacterial genomes, were
unidentifiable (denoted ``unknown''). This is common in viromes across
habitats; studies have reported as much as 95\% of virus sequences
belonging to unknown genomic units (14, 28--30). When the bacterial and
viral community signatures were combined, both bacterial and viral
organisms drove the community association with cancer \textbf{(Figure
\ref{predmodel} E)}.

@@

After evaluating our ability to classify samples between two disease
states, we performed a three-class random forest model including all
disease states. The 16S rRNA gene model yielded a mean AUC of 0.765 and
outperformed the viral community model, which yielded a mean AUC of
0.658 (p-value \textless{} 0.001, \textbf{Figure \ref{threewaymodel}
A-C}). The microbes important for the healthy versus cancer and healthy
versus adenoma models were also important for the three-class model
\textbf{(Figure \ref{threewaymodel} D-E)}. The most important bacterium
in the two and three class models was the same \emph{Fusobacterium} (OTU
4) \textbf{(Figure \ref{predmodel} C, Figure \ref{threewaymodel} D)}.
The viruses most important to the three-class model were identified as
bacteriophages \textbf{(Figure \ref{predmodel} D, Figure
\ref{threewaymodel} E)}, but not all important OVUs were of increased
abundance in the diseased state \textbf{(Figure \ref{threewaymodel} F)}.

>>>>

To identify the altered virus communities associated with colorectal
cancer, we built and tested random forest models for classifying stool
samples as belonging to individuals with either cancerous or healthy
colons. We confirmed that our bacterial 16S rRNA gene model replicated
the performance of the original report which used logit models instead
of random forest models \textbf{(Figure \ref{predmodel} A)} (8). We then
compared the bacterial OTU model to a model built using OVU relative
abundances. The viral model performed as well as the bacterial model
(corrected p-value = 0.6), with the viral and bacterial models achieving
mean area under the curve (AUC) values of 0.768 and 0.775, respectively
\textbf{(Figure \ref{predmodel} A - B)}. To evaluate the ability of both
bacterial and viral biomarkers to classify samples, we built a combined
model that used both bacterial and viral community data. The combined
model did not yield a statistically significant performance improvement
beyond the viral (corrected p-value = 0.08) and bacterial (corrected
p-value = 0.1) models, yielding an AUC of 0.807 \textbf{(Figure
\ref{predmodel} A - B)}.

We compared viral metagenomic methods to bacterial metagenomic methods
by building a viral model and a model built using OGU relative abundance
profiles from bacterial metagenomic shotgun sequencing data. This
bacterial model performed worse than the other models (mean AUC = 0.458)
\textbf{(Figure \ref{predmodel} A - B)}. To determine the cause of the
discrepancy between the two bacterial sequencing methods, we attempted
to compare the approaches at a common sequencing depth. This revealed
that the bacterial 16S rRNA gene model was strongly driven by sparse and
low abundance OTUs \textbf{(Figure \ref{16scompare})}. Removal of OTUs
with a median abundance of zero resulted in the removal of six OTUs, and
a loss of model performance down to what was observed in the
metagenome-based model \textbf{(Figure \ref{16scompare} A)}. The
majority of these OTUs had a relative abundance lower than 1\% across
the samples \textbf{(Figure \ref{16scompare} B)}. Although the features
in the viral model also were of low abundance \textbf{(Figure
\ref{threewaymodel} F)}, the coverage was sufficient for high model
performance, likely because viral genomes are orders of magnitude
smaller than bacterial genomes.

The association between the bacterial and viral communities and
colorectal cancer was driven by a few important microbes.
\emph{Fusobacterium} was the primary driver of the bacterial association
with colorectal cancer, which is consistent with its previously
described oncogenic potential \textbf{(Figure \ref{predmodel} C)}(22).
The virome signature also was driven by a few OVUs, suggesting a role
for these viruses in tumorigenesis \textbf{(Figure \ref{predmodel} D)}.
It is also important to note that while these viruses were driving the
signature, the magnitude of their importance and the significance of
those values was noticibly less than the bacterial 16S signature,
suggesting that unlike what is observed in the bacteria, there are many
viruses that are associated with the cancerous state. The identified
viruses were bacteriophages, belonging to \emph{Siphoviridae},
\emph{Myoviridae}, and phage taxa that could not be confidently
identified beyond their broad phage identification (i.e.
``unclassified''). Viruses, which were confirmed to not have genomic
similarity to known bacterial genomes, were unidentifiable (denoted
``unknown''). This is common in viromes across habitats; studies have
reported as much as 95\% of virus sequences belonging to unknown genomic
units (14, 28--30). When the bacterial and viral community signatures
were combined, both bacterial and viral organisms drove the community
association with cancer \textbf{(Figure \ref{predmodel} E)}.

@@

After evaluating our ability to classify samples between two disease
states, we performed a three-class random forest model including all
disease states. The 16S rRNA gene model yielded a mean AUC of 0.784 and
outperformed the viral community model, which yielded a mean AUC of
0.654 (p-value \textless{} 0.001, \textbf{Figure \ref{threewaymodel}
A-C}). The microbes important for the healthy versus cancer and healthy
versus adenoma models were also important for the three-class model
\textbf{(Figure \ref{threewaymodel} D-E)}. The most important bacterium
in the two and three class models was the same \emph{Fusobacterium} (OTU
4) \textbf{(Figure \ref{predmodel} C, Figure \ref{threewaymodel} D)}.
The viruses most important to the three-class model were identified as
bacteriophages \textbf{(Figure \ref{predmodel} D, Figure
\ref{threewaymodel} E)}, but not all important OVUs were of increased
abundance in the diseased state \textbf{(Figure \ref{threewaymodel} F)}.
